---
title: "zhkingProjectOne"
format: html
---
Loading packages:
```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(purrr)
library(readr)
```


# Part Zero: The Data

There's a lot of variables to work with here. We are specifically working with:

- AGEP : Age
- FER: Gave birth to a child in the past year
- HHL: Household Language
- GASP: Gas Cost
- GRPIP: Gross Rent as Percentage of Household Income
- JWAP: Time of Arrival at Work
- JWMNP: Travel Time to Work
- JWRIP: Vehicle Occupancy
- PWGTP: Person Weight
- SCH: School Enrollment
- SCHL: Educational Attainment
- SEX: Sex

And for geography:

- ST/STATE: State
- Region: Quadrant of the US
- Division: Region of the Quadrant of the US


# Part One: Data Processing

## Parsing Variable Values

Starting off, we need to save the links for the API call. 

```{r}

l21<-"https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
l22<-"https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
l23<-"https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
l24<-"https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

```
Next, we create a function that pulls our desired list of variables out of the API call, taking into account the ST/STATE difference between years.

```{r}

get_vars <- function(link){
  var_return <- httr::GET(link)
  var_parsed <- fromJSON(rawToChar(var_return$content))
  vars_llist <- var_parsed$variables
  vars_fin <- vars_llist[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE", "REGION", "DIVISION")]
  vars_fin[!sapply(vars_fin,is.null)]

}
```

Now we run this function on our links, and save all of the lists into one big list.

```{r}
var_list_21 <- get_vars(l21)
var_list_22 <- get_vars(l22)
var_list_23 <- get_vars(l23)
var_list_24 <- get_vars(l24)
var_lists <- list(var_list_21,var_list_22,var_list_23,var_list_24)
```

Finally, we save the data and remove extraneous R objects.

```{r}
saveRDS(var_lists, file = "data/var_lists.rds")
var_lists_check<-readRDS("data/var_lists.rds")
remove(l21,l22,l23,l24,get_vars,var_lists_check)
```


## Parsing API Data

## Combining Years


# Part Two: Writing a Generic Function for Summarizing

## Generic Functions

## Numerical Summaries

## Graphical Summaries 

# Investigation