---
title: "zhkingProjectOne"
format: html
---
Loading packages:
```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(purrr)
library(readr)
library(lubridate)
```


# Part Zero: The Data

There's a lot of variables to work with here. We are specifically working with:

- AGEP : Age
- FER: Gave birth to a child in the past year
- HHL: Household Language
- GASP: Gas Cost
- GRPIP: Gross Rent as Percentage of Household Income
- JWAP: Time of Arrival at Work
- JWMNP: Travel Time to Work
- JWRIP: Vehicle Occupancy
- PWGTP: Person Weight
- SCH: School Enrollment
- SCHL: Educational Attainment
- SEX: Sex

And for geography:

- ST/STATE: State
- Region: Quadrant of the US
- Division: Region of the Quadrant of the US


# Part One: Data Processing

## Parsing Variable Values

Starting off, we need to save the links for the API call. 

```{r}

l21<-"https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
l22<-"https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
l23<-"https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
l24<-"https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

```
Next, we create a function that pulls our desired list of variables out of the API call, taking into account the ST/STATE difference between years.

```{r}

get_vars <- function(link){
  var_return <- httr::GET(link)
  var_parsed <- fromJSON(rawToChar(var_return$content))
  vars_llist <- var_parsed$variables
  vars_fin <- vars_llist[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE", "REGION", "DIVISION")]
  vars_fin[!sapply(vars_fin,is.null)]

}
```

Now we run this function on our links, and save all of the lists into one big list.

```{r}
var_list_21 <- get_vars(l21)
var_list_22 <- get_vars(l22)
var_list_23 <- get_vars(l23)
var_list_24 <- get_vars(l24)
var_lists <- list("y2021"=var_list_21,"y2022"=var_list_22,"y2023"=var_list_23,"y2024"=var_list_24)
```

Finally, we save the data and remove extraneous R objects.

```{r}
saveRDS(var_lists, file = "data/var_lists.rds")
var_lists_check<-readRDS("data/var_lists.rds")
remove(l21,l22,l23,l24,get_vars,var_lists_check)
```


## Parsing API Data

First, we'll do a test process using the example URL from the rubric.

```{r}

censusKey<-readRDS("../censusKey.rds")
base_url <- "https://api.census.gov/data/2024/acs/acs1/pums"

api_ex <- GET(
  url = base_url,
  query = list(
    get = "SEX,PWGTP,JWAP",
    `for` = "state:*",
    SCHL = "24",
    key = censusKey
  )
)
raw_stats <- content(api_ex, "text", encoding = "UTF-8")
parsed_stats <- fromJSON(raw_stats)
example_stats <- as_tibble(parsed_stats)
colnames(example_stats) <- example_stats[1, ]
example_stats <- example_stats[-1, , drop = FALSE]

remove(api_ex,raw_stats,parsed_stats)
```

A short aside: we need the list of geography codes, so I'll make a helper function to do that for me.

```{r}
grabCodes<-function(){
  check<-var_lists[[1]]
  regions<-check$REGION
  values<-regions$values
  regionCodes<-values$item
  states<-check$ST
  values<-states$values
  stateCodes<-values$item
  divisions<-check$DIVISION
  values<-divisions$values
  divisionCodes<-values$item
  codes<-list("regionCodes"=names(regionCodes),"stateCodes"=names(stateCodes),"divisionCodes"=names(divisionCodes))
  return(codes)
}
codes<-grabCodes()
```

We also need a helper function to turn the times from JWAP into the midpoints. As mentioned in the rubric, I did use a LLm for this.

```{r}
get_time_midpoint <- function(time_ranges, format_12hr = TRUE) {
  # 1. Standardize "p.m." -> "PM" and "a.m." -> "AM"
  cleaned <- gsub("p\\.m\\.", "PM", time_ranges, ignore.case = TRUE)
  cleaned <- gsub("a\\.m\\.", "AM", cleaned, ignore.case = TRUE)
  
  # 2. Split start and end
  split_times <- strsplit(cleaned, "\\s+to\\s+")
  start_str <- sapply(split_times, `[`, 1)
  end_str   <- sapply(split_times, `[`, 2)
  
  # 3. Parse with anchored dummy date
  today <- Sys.Date()
  start_time <- as.POSIXct(paste(today, start_str), format = "%Y-%m-%d %I:%M %p")
  end_time   <- as.POSIXct(paste(today, end_str),   format = "%Y-%m-%d %I:%M %p")
  
  # 4. Handle overnight boundary crossover
  overnight <- !is.na(start_time) & !is.na(end_time) & (end_time < start_time)
  end_time[overnight] <- end_time[overnight] + (24 * 3600)
  
  # 5. Compute midpoint
  midpoint <- start_time + (end_time - start_time) / 2
  
  # 6. Format output (12-hour AM/PM vs 24-hour)
  if (format_12hr) {
    return(format(midpoint, "%I:%M %p")) # Returns e.g. "06:37 PM"
  } else {
    return(format(midpoint, "%H:%M"))    # Returns e.g. "18:37"
  }
}

# Now we want to apply this to JWAP potential values.

get_JWAPPED<-function(){
  check<-var_lists[[1]]
  jwap1<-check$JWAP
  values<-jwap1$values
  JWAP<-unlist(values$item)
  ids<-names(JWAP)
  JWAP<-cbind("JWAP"=ids,"times"=JWAP)
  JWAP[,2]<-get_time_midpoint(JWAP[,2])
  JWAP<-as_tibble(JWAP)
  return(JWAP)
}

get_JWDPPED<-function(){
  check<-var_lists[[1]]
  jwdp1<-check$JWDP
  values<-jwdp1$values
  JWDP<-unlist(values$item)
  ids<-names(JWDP)
  JWDP<-cbind("JWDP"=ids,"times"=JWDP)
  JWDP[,2]<-get_time_midpoint(JWDP[,2])
  JWDP<-as_tibble(JWDP)
  return(JWDP)
}

JWAPvals<-get_JWAPPED()
JWDPvals<-get_JWDPPED()

```


Next, we'll turn the example into a helper function that can do the same thing for any call.

```{r}

api_helper <- function(KEY,YEAR,AGEP=TRUE,GRPIP=FALSE,JWAP=FALSE,JWDP=FALSE,JWMNP=FALSE, 
                       SEX=TRUE, HHL=FALSE, SCH=FALSE, SCHL=FALSE, GEOG="ST",GSUBSET=99999){
  
  # First Step is to make sure that all of the inputs are correct.
  
  if (!is.numeric(GSUBSET) || 
      !all(is.logical(c(AGEP, GRPIP, JWAP, JWDP, JWMNP, SEX, HHL, SCH, SCHL))) || 
      !all(is.character(c(GEOG, KEY)))) {
    stop("Wrong Kind of Input")
  }
  if(!YEAR %in% c(2021,2022,2023,2024)) {
    stop("Year Not in Range")}
  if(!GEOG %in% c("ST","STATE","REGION","DIVISION")){
    stop("Not a Geography Option")}
  
  # Next, I need to fix the whole ST/STATE issue.
  
  if(!YEAR %in% c(2023,2024) & GEOG == "ST"){
    GEOG <- "STATE"}
  if(!YEAR %in% c(2021,2022) & GEOG == "STATE"){
    GEOG <- "ST"}
  
  # Here, I'm checking to make sure all of the geography sub-setting lines up.
  
  GSUBSET <- as.character(GSUBSET)
  if(GEOG %in% c("ST","STATE") & !GSUBSET %in% codes$stateCodes){
    GSUBSET<-"01"}
  if(GEOG %in% c("REGION") & !GSUBSET %in% codes$regionCodes){
    GSUBSET<-"1"}
  if(GEOG %in% c("DIVISION") & !GSUBSET %in% codes$divisionCodes){
    GSUBSET<-"1"}
  
  # This is the baseline URL without anything applied yet.

  baseURL <- paste0("https://api.census.gov/data/",as.character(YEAR),"/acs/acs1/pums")
  
  # Now I'm setting up the default variables, and adding the selected variables to the pile.
  
  baseVars <- c("PWGTP","GASP","FER")
  specVars <-c(AGEP = "AGEP", GRPIP = "GRPIP", JWAP = "JWAP", JWDP = "JWDP", JWMNP = "JWMNP", 
               SEX = "SEX", HHL="HHL", SCH="SCH", SCHL="SCHL")
  selectedSpecVars <- specVars[c(AGEP, GRPIP, JWAP, JWDP, JWMNP,SEX,HHL,SCH,SCHL)]
  realVars <- paste(c(baseVars, selectedSpecVars), collapse = ",")
  
  # Here's where I'm making the geography line:
  
  realGeog <- paste0(tolower(GEOG),":",GSUBSET)
  
  # We should now finally be ready to actually call the API
  
  request <- GET(
    url = baseURL,
    query = list(
      get = realVars,
      `for` = realGeog,
      key = KEY
    )
  )
  raw_stats <- content(request, "text", encoding = "UTF-8")
  parsed_stats <- fromJSON(raw_stats)
  un_stats <- as_tibble(parsed_stats)
  colnames(un_stats) <- un_stats[1, ]
  un_stats <- un_stats[-1, , drop = FALSE]
  un_stats<-un_stats |>
    mutate(across(c(PWGTP, GASP, AGEP, GRPIP,JWMNP), as.double))|>
    mutate(across(c(FER,SEX,HHL,SCH,SCHL,JWAP,state), as.factor))|>
    left_join(JWAPvals, by = "JWAP") |>
    mutate(JWAP = times)|>
    select(-times)|>
    left_join(JWDPvals, by = "JWDP") |>
    mutate(JWDP = times)|>
    select(-times)
  
}




```


## Combining Years


# Part Two: Writing a Generic Function for Summarizing

## Generic Functions

## Numerical Summaries

## Graphical Summaries 

# Investigation