---
title: "zhkingProjectOne"
format: html
---
Loading packages:
```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(purrr)
```


# Part Zero: The Data

There's a lot of variables to work with here. We are specifically working with:

- AGEP : Age
- FER: Gave birth to a child in the past year
- HHL: Household Language
- GASP: Gas Cost
- GRPIP: Gross Rent as Percentage of Household Income
- JWAP: Time of Arrival at Work
- JWMNP: Travel Time to Work
- JWRIP: Vehicle Occupancy
- PWGTP: Person Weight
- SCH: School Enrollment
- SCHL: Educational Attainment
- SEX: Sex

And for geography:

- ST/STATE: State
- Region: Quadrant of the US
- Division: Region of the Quadrant of the US


# Part One: Data Processing

## Parsing Variable Values

We start by pulling a list of variables from the API, for each year. 

```{r}

v21 <- "https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
var21_return <- httr::GET(v21)
v21_parsed <- fromJSON(rawToChar(var21_return$content))
remove(v21,var21_return)
var21<-v21_parsed$variables
var21<-var21[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]

```

```{r}
get_vars <- function(link){
  var_return <- httr::GET(link)
  var_parsed <- fromJSON(rawToChar(var_return$content))
  vars_llist <- var_parsed$variables
  vars_fin <- vars_llist[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE", "REGION", "DIVISION")]
}



```



## Parsing API Data

## Combining Years


# Part Two: Writing a Generic Function for Summarizing

## Generic Functions

## Numerical Summaries

## Graphical Summaries 

# Investigation