---
title: "zhkingProjectOne"
format: html
---
Loading packages:
```{r}
#| echo: false
library(tidyverse)
library(httr)
library(jsonlite)
library(purrr)
library(readr)
library(lubridate)
library(stringr)
```


# Part Zero: The Data

There's a lot of variables to work with here. We are specifically working with:

- AGEP : Age
- FER: Gave birth to a child in the past year
- HHL: Household Language
- GASP: Gas Cost
- GRPIP: Gross Rent as Percentage of Household Income
- JWAP: Time of Arrival at Work
- JWMNP: Travel Time to Work
- JWRIP: Vehicle Occupancy
- PWGTP: Person Weight
- SCH: School Enrollment
- SCHL: Educational Attainment
- SEX: Sex

And for geography:

- ST/STATE: State
- Region: Quadrant of the US
- Division: Region of the Quadrant of the US


# Part One: Data Processing

## Parsing Variable Values

Starting off, we need to save the links for the API call. 

```{r}

l21<-"https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
l22<-"https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
l23<-"https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
l24<-"https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

```

Next, we create a function that pulls our desired list of variables out of the API call, taking into account the ST/STATE difference between years.

```{r}

getVars <- function(link){
  varReturn <- httr::GET(link)
  varParsed <- fromJSON(rawToChar(varReturn$content))
  varLlist <- varParsed$variables
  targetVars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE", "REGION", "DIVISION")
  availableVars <- intersect(targetVars, names(varLlist))
  varsFin <- varLlist[availableVars]
  
  return(varsFin[!sapply(varsFin, is.null)])
}
```

Now we run this function on our links, and save all of the lists into one big list.

```{r}
varList21 <- getVars(l21)
varList22 <- getVars(l22)
varList23 <- getVars(l23)
varList24 <- getVars(l24)
varLists <- list("y2021"=varList21,"y2022"=varList22,"y2023"=varList23,"y2024"=varList24)
```

Finally, we save the data and remove extraneous R objects.

```{r}
saveRDS(varLists, file="data/varLists.rds")
remove(l21,l22,l23,l24,getVars)
```


## Parsing API Data

First, we'll do a test process using the example URL from the rubric.

```{r}

censusKey<-readRDS("../censusKey.rds")
baseUrl <- "https://api.census.gov/data/2024/acs/acs1/pums"

apiEx <- GET(
  url=baseUrl,
  query=list(
    get="SEX,PWGTP,JWAP",
    `for`="state:01,02",
    SCHL="24",
    key=censusKey
  )
)
rawStats <- content(apiEx, "text", encoding="UTF-8")
parsedStats <- fromJSON(rawStats)
exampleStats <- as_tibble(parsedStats)
colnames(exampleStats) <- exampleStats[1, ]
exampleStats <- exampleStats[-1, , drop=FALSE]

remove(base,url,apiEx,rawStats,parsedStats,exampleStats)
```

We need a bunch of helper functions before we can get to the real deal.

First, we need a function that checks if the basic variable inputs make sense.

```{r}
matchInputsSimple <- function(key,year,agep,gasp,grpip,JWAP,JWDP,jwmnp,sex,fer,hhl,sch,schl,geog,gsubset){
  if(!all(is.logical(c(agep,gasp,grpip,JWAP,JWDP,jwmnp,sex,fer,hhl,sch,schl)))){
    stop("Each of the non-filtering variables takes a T/F value")
  }
  if(!is.character(key)){
    stop("The census key must be entered as a string")
  }
  geog <- tolower(geog)
  if(!geog %in% c("state","region","division")){
    stop("The geography options are state, region, and division.")
  }
}
```

Next, we need to check if the user has selected at least one numeric and one categorical variable.

```{r}
atLeastOne <- function(numVars,catVars){
  if (!any(numVars)) {
    stop("Choose at least one numeric variable.")}
  if (!any(catVars)) {
    stop("Choose at least one categorical variable.")
  }
}
```

Now we'll create the string for the selected (and prescribed) variables.

```{r}
setVarString <- function(numVars,catVars){
  selectedNums <- names(numVars)[numVars]
  selectedCats <- names(catVars)[catVars]
  allRequested <- unique(c("PWGTP", selectedNums, selectedCats))
  realVars <- toupper(paste(allRequested, collapse=","))
}
```

Now we need to deal with the geography. The rubric implies a difference between the years 2021/2022 and 2023/2024 in terms of the st/state variable, however when I attempted to account for this I discovered that it was not the case on my end. If you go back through my commits you can see when I figured this out.

Anyway, let's start by getting a list of the codes for the geography levels.

```{r}

grabCodes<-function(){
  check<-varLists[[1]]
  regions<-check$REGION
  values<-regions$values
  regionCodes<-values$item
  states<-check$ST
  values<-states$values
  stateCodes<-values$item
  divisions<-check$DIVISION
  values<-divisions$values
  divisionCodes<-values$item
  geogCodes<-list("regionCodes"=names(regionCodes),"stateCodes"=names(stateCodes),"divisionCodes"=names(divisionCodes))
  return(geogCodes)
}
geogCodes <- grabCodes()

```

Moving on, let's check to make sure that the inputted geography level matches the setting.

```{r}
matchGeogSubsets <- function(geog,gsubset){if(geog == "region"){
  options<-geogCodes$regionCodes
  options[[6]]<-"All"}
  if(geog == "division"){
    options<-geogCodes$divisionCodes
    options[[11]]<-"All"}
  if(geog == "state"){
    options<-geogCodes$stateCodes
    options[[53]]<-"All"}
  
  if(!gsubset %in% options){
    stop(paste0("The ", geog," code selected does not exist, choose from: ",paste(options,collapse=",")))}}
```

Now we need to account for the "all" option. I think you can do this with the asterisk, but in my fevered attempts to fix the st/state issue, I changed it to this method, and it works, so I'm not changing it again.

```{r}
allowForAll <- function(geog,gsubset){
  if (gsubset == "All"){
    if(geog == "region"){gsubset <-allRegions}
    else if (geog == "division"){gsubset <- allDivisions}
    else if (geog == "state"){gsubset <- allStates}
  }
  else {gsubset <- gsubset}
}
```

Now we'll set up the string for the geography settings.

```{r}
setGeogString <- function(geog,gsubset){
  geogString <- paste0(geog,":",gsubset)
}
```

Now I need to do some final set-up for fixing up the final tibble. 

First, we'll ask AI to help us with the midpoint function:

```{r}
get_time_midpoint <- function(time_ranges, format_12hr=TRUE) {
  # 1. Standardize "p.m." -> "PM" and "a.m." -> "AM"
  cleaned <- gsub("p\\.m\\.", "PM", time_ranges, ignore.case=TRUE)
  cleaned <- gsub("a\\.m\\.", "AM", cleaned, ignore.case=TRUE)
  
  # 2. Split start and end
  split_times <- strsplit(cleaned, "\\s+to\\s+")
  start_str <- sapply(split_times, `[`, 1)
  end_str   <- sapply(split_times, `[`, 2)
  
  # 3. Parse with anchored dummy date
  today <- Sys.Date()
  start_time <- as.POSIXct(paste(today, start_str), format="%Y-%m-%d %I:%M %p")
  end_time   <- as.POSIXct(paste(today, end_str),   format="%Y-%m-%d %I:%M %p")
  
  # 4. Handle overnight boundary crossover
  overnight <- !is.na(start_time) & !is.na(end_time) & (end_time < start_time)
  end_time[overnight] <- end_time[overnight] + (24 * 3600)
  
  # 5. Compute midpoint
  midpoint <- start_time + (end_time - start_time) / 2
  
  # 6. Format output (12-hour AM/PM vs 24-hour)
  if (format_12hr) {
    return(format(midpoint, "%I:%M %p")) # Returns e.g. "06:37 PM"
  } else {
    return(format(midpoint, "%H:%M"))    # Returns e.g. "18:37"
  }
}
```

And now we apply this to the values for JWAP and JWDP, which are unfortunately not the same.

```{r}
getJWAPPED<-function(){
  check<-varLists[[1]]
  JWAP1<-check$JWAP
  values<-JWAP1$values
  JWAP<-unlist(values$item)
  ids<-names(JWAP)
  JWAP<-cbind("JWAP"=ids,"times"=JWAP)
  JWAP[,2]<-get_time_midpoint(JWAP[,2])
  JWAP<-as_tibble(JWAP)
  return(JWAP)
}

getJWDPPED<-function(){
  check<-varLists[[1]]
  JWDP1<-check$JWDP
  values<-JWDP1$values
  JWDP<-unlist(values$item)
  ids<-names(JWDP)
  JWDP<-cbind("JWDP"=ids,"times"=JWDP)
  JWDP[,2]<-get_time_midpoint(JWDP[,2])
  JWDP<-as_tibble(JWDP)
  return(JWDP)
}

JWAPVals<-getJWAPPED()
JWDPVals<-getJWDPPED()
```

Finally, we pull on all these functions to create the opposite of a helper function: a harmer function.

```{r} 
apiHarmer <- function(key,year=2024,agep=T,gasp=F,grpip=F,JWAP=F,JWDP=F,jwmnp=F,sex=T,fer=F,hhl=F,sch=F,schl=F,geog="region",gsubset="2"){
  
  geogCodes <- grabCodes()
  JWAPVals<-getJWAPPED()
  JWDPVals<-getJWDPPED()
  
  allStates <- paste(geogCodes$stateCodes, collapse=",")
  allRegions <- paste(geogCodes$regionCodes, collapse=",")
  allDivisions <- paste(geogCodes$divisionCodes, collapse=",")
  
  numVars <- c(agep=agep, gasp=gasp,grpip=grpip,JWAP=JWAP,JWDP=JWDP,jwmnp=jwmnp)
  catVars <- c(sex=sex,fer=fer,hhl=hhl,sch=sch,schl=schl)
  
  matchInputsSimple(key,year,agep,gasp,grpip,JWAP,JWDP,jwmnp,sex,fer,hhl,sch,schl,geog,gsubset)
  
  atLeastOne(numVars,catVars)
  
  varString <- setVarString(numVars,catVars)
  
  matchGeogSubsets(geog,gsubset)
  
  gsubset<-allowForAll(geog,gsubset)
  
  geogString <- setGeogString(geog,gsubset)
  
  censusUrl <- paste0(
    "https://api.census.gov/data/", year, "/acs/acs1/pums",
    "?get=", varString,
    "&for=", geogString,
    "&key=", key
  )
  
  request <- GET(
    url=censusUrl
  )
  rawStats <- content(request, "text", encoding="UTF-8")
  parsedStats <- fromJSON(rawStats)
  censusPull <- as_tibble(parsedStats)
  colnames(censusPull) <- censusPull[1, ]
  censusPull <- censusPull[-1, , drop=FALSE]
  
  # Here I'm joining the output with my JWAPVals tibble, so that I can swap out the time codes for the actual times
  if ("JWAP" %in% names(censusPull)) {
    censusPull <- censusPull |>
      left_join(JWAPVals, by="JWAP") |> 
      mutate(JWAP=times) |> # here's where I do that swap
      select(-times)}    # now I'm removing the leftovers of the join
  
  # and now I repeat the process for JWDP, which unfortunately has different values from JWAP
  if ("JWDP" %in% names(censusPull)) {
    censusPull <- censusPull |>
      left_join(JWDPVals, by="JWDP") |>
      mutate(JWDP=times) |>
      select(-times)}
  
  censusPull <- censusPull |>
    mutate(across(any_of(c("PWGTP", "GASP", "AGEP", "GRPIP", "JWMNP")), as.double))
  
  yearKey <- paste0("y",year)
  yearMeta <- varLists[[yearKey]]
  
  selectedCats <- names(catVars)[catVars]
  
  for (cat_var in selectedCats) {
    cat_upper <- toupper(cat_var)
    
    meta_key <- if (cat_upper %in% names(yearMeta)) cat_upper else if (cat_var %in% names(yearMeta)) cat_var else NULL
    
    if (cat_upper %in% names(censusPull) && !is.null(meta_key)) {
      valItems <- yearMeta[[meta_key]]$values$item
      if (!is.null(valItems)) {
        levelsVec <- as.character(names(valItems))
        labelsVec <- unname(unlist(valItems))
        
        censusPull <- censusPull |>
          mutate(!!cat_upper := factor(.data[[cat_upper]], levels=levelsVec, labels=labelsVec))
      }
    }
  }
  
  class(censusPull) <- c("census", class(censusPull))
  
  return(censusPull)
  
}
```


## Combining Years

This helper function allows for us to call multiple years, using the original apiHarmer function, and a built-in helper function this time.

```{r}

censusAPIMultiyear <- function(key,
                               Y2021=FALSE, Y2022=FALSE, Y2023=FALSE, Y2024=TRUE, 
                               agep=TRUE,
                               gasp=FALSE, grpip=FALSE, JWAP=FALSE, JWDP=FALSE, jwmnp=FALSE, 
                               sex=TRUE, 
                               fer=FALSE, hhl=FALSE, sch=FALSE, schl=FALSE, 
                               geog="STATE", gsubset="01") {
  
  puller <- function(yearN) {
    block <- apiHarmer(
      key=key, 
      year=yearN, 
      agep=agep, gasp=gasp, grpip=grpip, JWAP=JWAP, JWDP=JWDP, jwmnp=jwmnp,
      sex=sex, fer=fer, hhl=hhl, sch=sch, schl=schl,
      geog=geog, gsubset=gsubset
    )
    block$Year <- yearN
    return(block)
  }
  
  dataList <- list()
  
  if (Y2021) { dataList[["2021"]] <- puller(2021) }
  if (Y2022) { dataList[["2022"]] <- puller(2022) }
  if (Y2023) { dataList[["2023"]] <- puller(2023) }
  if (Y2024) { dataList[["2024"]] <- puller(2024) }
  
  if (length(dataList) == 0) {
    stop("Choose at least 1 year.")
  }
  
  comboYears <- bind_rows(dataList)
  
  class(comboYears) <- c("census", class(comboYears))
  return(comboYears)
}
```


# Part Two: Writing a Generic Function for Summarizing

## Numerical Summaries

Moving on from the API Call itself, it's time to create our summary function.


```{r}
summary.census <- function(census,
                           numerics=c("AGEP","GASP","GRPIP","JWMNP"),
                           categorics=c("SEX","FER","HHL","SCH","SCHL")){
  
    # First, let's check to make sure all of our inputs match again.
  
    matchInputs2 <- function(census,numerics,categorics){
    if(!class(census)[1] == "census"){
      stop("Census object must be of the Census class.")
    }
    if(!is.atomic(numerics)){
      stop("The numeric variables must be in a atomic vector")
    }
    if(!is.atomic(categorics)){
      stop("The categorical variables must be in a atomic vector")
    }
    if (!all(sapply(numerics, is.character))){
      stop("The items of the numeric variable list must be the string names of the variables.")
    }
    if (!all(sapply(categorics, is.character))){
      stop("The items of the cateogrical variable list must be the string names of the variables.")}}
    
    # here's where we actually run that
    matchInputs2(census,numerics,categorics)
    
    # Next up, we'll check the selected variables against the census variables and the ones we can actually summarize here (no times or pgtwp)
    varsInCensus <- function(census,numerics,categorics){
      potVars <- colnames(census)
      allPotVars <- c("AGEP","GASP","GRPIP","JWMNP","SEX","FER","HHL","SCH","SCHL")
      potVars <- potVars[potVars %in% allPotVars]
      numerics <- numerics[numerics %in% potVars]
      categorics <- categorics[categorics %in% potVars]
      if(is_empty(numerics) & is_empty(categorics)){
        stop("No valid variables selected")}
      endVars <- list("num"=numerics,"cat"=categorics)
      return(endVars)}
  
    # here's where we actually run that helper function
    endVars<-varsInCensus(census,numerics,categorics)  
    numerics<-endVars$num
    categorics<-endVars$cat
    
    # Now I'm splitting the census into the numeric and categorical variable parts.
    numCensus <- select(census,all_of(numerics))
    catCensus <- select(census,all_of(categorics))
    
    # From that, I'm using sapply to get the mean and sd of every selected numerical variable.
    numSum <- data.frame(
      variable=names(numCensus),
      mean=sapply(numCensus, mean, na.rm=TRUE),
      sd=sapply(numCensus, sd, na.rm=TRUE),
      row.names=NULL)
    
    # This was the worst part. There is probably a more efficient way of doing this, but this is what I came up with     after a lot of googling.
    catSum <- do.call(rbind, lapply(names(catCensus), function(var) {
                              counts <- table(catCensus[[var]], useNA="ifany")
                              data.frame(variable=var,level=names(counts),count=as.vector(counts),row.names=NULL)}))
    
    # Now we combine those two into a final list.
    finSum <- list("Numerical Variables" = numSum,"Categorical Variables" = catSum)
    return(finSum)
}

```


## Graphical Summaries 

The final code writing section: creating a generic plot function. 

```{r}
plot.census <- function ()
```


# Investigation