---
title: "zhkingProjectOne"
format: html
---
Loading packages:
```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(purrr)
library(readr)
```


# Part Zero: The Data

There's a lot of variables to work with here. We are specifically working with:

- AGEP : Age
- FER: Gave birth to a child in the past year
- HHL: Household Language
- GASP: Gas Cost
- GRPIP: Gross Rent as Percentage of Household Income
- JWAP: Time of Arrival at Work
- JWMNP: Travel Time to Work
- JWRIP: Vehicle Occupancy
- PWGTP: Person Weight
- SCH: School Enrollment
- SCHL: Educational Attainment
- SEX: Sex

And for geography:

- ST/STATE: State
- Region: Quadrant of the US
- Division: Region of the Quadrant of the US


# Part One: Data Processing

## Parsing Variable Values

Starting off, we need to save the links for the API call. 

```{r}

l21<-"https://api.census.gov/data/2021/acs/acs1/pums/variables.json"
l22<-"https://api.census.gov/data/2022/acs/acs1/pums/variables.json"
l23<-"https://api.census.gov/data/2023/acs/acs1/pums/variables.json"
l24<-"https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

```
Next, we create a function that pulls our desired list of variables out of the API call, taking into account the ST/STATE difference between years.

```{r}

get_vars <- function(link){
  var_return <- httr::GET(link)
  var_parsed <- fromJSON(rawToChar(var_return$content))
  vars_llist <- var_parsed$variables
  vars_fin <- vars_llist[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE", "REGION", "DIVISION")]
  vars_fin[!sapply(vars_fin,is.null)]

}
```

Now we run this function on our links, and save all of the lists into one big list.

```{r}
var_list_21 <- get_vars(l21)
var_list_22 <- get_vars(l22)
var_list_23 <- get_vars(l23)
var_list_24 <- get_vars(l24)
var_lists <- list("y2021"=var_list_21,"y2022"=var_list_22,"y2023"=var_list_23,"y2024"=var_list_24)
```

Finally, we save the data and remove extraneous R objects.

```{r}
saveRDS(var_lists, file = "data/var_lists.rds")
var_lists_check<-readRDS("data/var_lists.rds")
remove(l21,l22,l23,l24,get_vars,var_lists_check)
```


## Parsing API Data

First, we'll do a test process using the example URL from the rubric.

```{r}

census_key<-readRDS("../censusKey.rds")
base_url <- "https://api.census.gov/data/2024/acs/acs1/pums"

api_ex <- GET(
  url = base_url,
  query = list(
    get = "SEX,PWGTP,MAR",
    `for` = "state:*",
    SCHL = "24",
    key = census_key
  )
)
raw_stats <- content(api_ex, "text", encoding = "UTF-8")
parsed_stats <- fromJSON(raw_stats)
example_stats <- as_tibble(parsed_stats[-1, , drop = FALSE])
remove(api_ex,raw_stats,parsed_stats)



```

A short aside: we need the list of geography codes, so I'll make a helper function to do that for me.

```{r}
grabCodes<-function(){
  check<-var_lists[[1]]
  regions<-check$REGION
  values<-regions$values
  regionCodes<-values$item
  states<-check$ST
  values<-states$values
  stateCodes<-values$item
  divisions<-check$DIVISION
  values<-divisions$values
  divisionCodes<-values$item
  codes<-list("regionCodes"=regionCodes,"stateCodes"=stateCodes,"divisionCodes"=divisionCodes)
  return(codes)
}
codes<-grabCodes()
```


Next, we'll turn the example into a helper function that can do the same thing for any call.

```{r}

api_helper <- function(YEAR,AGEP=TRUE,GRPIP=FALSE,JWAP=FALSE,JWDP=FALSE,JWMNP=FALSE){
  if(!is.integer(vec) | !is.logical(c(AGEP,GRPIP,JWAP,JWDP,JWMNP))){
    stop("Wrong Kind of Input")
  }
  if(!YEAR %in% c(2021,2022,2023,2024)) {
    stop("Year Not in Range")
  }
  baseURL <- paste0("https://api.census.gov/data/",as.character(YEAR),"/acs/acs1/pums")
  baseVARS <- "PWGTP,GASP"
  ifelse(AGEP==TRUE,baseVARS<-paste0(baseVARS,""))
  
  request <- GET(
  url = base_url,
  query = list(
    get = "SEX,PWGTP,MAR",
    `for` = "state:*",
    SCHL = "24",
    key = census_key
  )
)
  
}



```


## Combining Years


# Part Two: Writing a Generic Function for Summarizing

## Generic Functions

## Numerical Summaries

## Graphical Summaries 

# Investigation